---
title: "Assignment3ExtraCredit"
author: "Matt Lewitzky"
date: "2022-09-21"
output: html_document
---

16.  Find your own dataset, import it and implement the following functions on the data

- head
- str
- names
- mean, min, max
- table
- is.na
- colSums
- class
- cor
- by
- ifelse
- case_when

```{r}
setwd("C:/Users/student/OneDrive - Bryant University/Desktop/Math 421")
df <- read.csv('adult_census.csv')
head(df)
```
```{r}
str(df)
```
```{r}
names(df)
```
```{r}
mean(df$hours.per.week, na.rm=TRUE)
max(df$hours.per.week, na.rm=TRUE)
min(df$hours.per.week, na.rm=TRUE)
```
```{r}
table(df$sex)
```
```{r}
sum(is.na(df))
```
```{r}
colSums(is.na(df))
```
```{r}
class(df$sex)
```
```{r}
cor(df$hours.per.week, df$age)
```
```{r}
by(df$age, df$hours.per.week, FUN = mean)
```
```{r}
df$sex2 <- ifelse(df$sex=='Female', 
                        0,1)
df
```
```{r}
library(dplyr)
df$age2 <-case_when(df$age<25 ~ 'Young Person',
                    df$age<60 ~ 'Middle Aged',
                    TRUE ~ 'Older')
head(df)
```
17.  In the dataset in #16, practice the follows.  You can reuse the code of 16. 

- Create a categorical variable from a continuous variable
- From one categorical variable, create a new categorical variable with fewer categories
```{r}
df$age2 <- case_when(df$age<12 ~ 'child',
                             df$age<20 ~ 'teenager', 
                            df$age<40 ~ 'middle ager',
                              df$age<70 ~ 'old',
                             TRUE~'elderly')
table(df$age2)
```

```{r}
df$age3 <- case_when(df$age2 %in% c('child','teenager') ~ 'younger',
                       df$age2 %in% c('middle ager','old','elderly') ~ 'older',
                       TRUE~'break')
table(df$age3)

```



