
---
output: 
  html_document:
  pdf_document: default
  word_document: default
title: "Assignment 7 - Extra Credits: Practice dplyr and ggplot2"
---

***How to do it?***: 

- Open the Rmarkdown file of this assignment ([link](assignment7_extra_credits.Rmd)) in Rstudio. 

- Right under each **question**, insert  a code chunk (you can use the hotkey `Ctrl + Alt + I` to add a code chunk) and code the solution for the question. 

- `Knit` the rmarkdown file (hotkey: `Ctrl + Alt + K`) to export an html.  

-  Publish the html file to your Githiub Page. 

***Submission***: Submit the link on Github of the assignment to Canvas



```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```


-------

1. Import the data at https://bryantstats.github.io/math421/data/ds_salaries.csv

    The data description can be found at https://www.kaggle.com/datasets/saurabhshahane/data-science-jobs-salaries
```{r}
library(tidyverse)
df <- read_csv('C:/Users/student/OneDrive - Bryant University/Desktop/Math 421/ds_salaries.csv')

head(df)
```

2. Is there any column that contains NAs?
```{r}
colSums(is.na(df))
```

3. What is the most frequent job title in the data?  
```{r}
names(which.max(table(df$job_title)))
```

4. Find the average salary data of the three most frequent job titles in the data. 
```{r}
df %>% count(job_title) %>% arrange(-n)
```
```{r}
df %>% 
  filter(job_title %in% c('Data Scientist', 'Data Engineer', 'Data Analyst')) %>%
  summarize(avg_salary = mean(salary_in_usd))
```

5. What is the job title that have the highest average salary?
```{r}
df %>% 
  group_by(job_title) %>%
  summarize(avg_salary = mean(salary_in_usd)) %>%
  arrange(-avg_salary)
```

6. Which country gives the highest salary on average? 
```{r}
df %>% 
  group_by(company_location) %>%
  summarize(avg_salary = mean(salary_in_usd)) %>%
  arrange(-avg_salary)
```

7. Plot five meaningful plots. All plots should have title, caption, and appropriate labels.
```{r}
df %>% 
  filter(job_title %in% c('Data Analyst', 'Data Scientist', 'ML Engineer')) %>%
  group_by(job_title) %>%
  summarize(avg_salary = mean(salary_in_usd)) %>% 
  ggplot()+ 
  geom_point(mapping=aes(x=job_title, y=avg_salary))+
  labs(x='Job Title', 
       y = 'Average Salary', 
       title = 'Average Salaries per Job', 
       caption = 'ML Engineers have the best salary!')
```
```{r}
df %>% 
  filter(job_title %in% c('Data Analyst', 'Data Scientist', 'ML Engineer')) %>%
  group_by(company_size) %>%
  summarize(avg_salary = mean(salary_in_usd)) %>% 
  ggplot()+ 
  geom_point(mapping=aes(x=company_size, y=avg_salary))+
  labs(x='Company Size', 
       y = 'Average Salary', 
       title = 'Average Salaries by Company Size', 
       caption = 'The medium sized companies have the best salaries!')
```
```{r}
df %>%
  group_by(company_size, experience_level) %>% 
  summarize(avg_salary = mean(salary_in_usd)) %>%
  ggplot()+ 
  geom_col(mapping=aes(x=company_size, y=avg_salary, fill=experience_level))+
  labs(x='Company Size', 
       y = 'Average Salary',
       color = 'Experience Level',
       title = 'Average Salary by the Company Size and the Experience Level', 
       caption = 'This graph represents the average salary by company size and experience level.')
```
```{r}
df %>%
  group_by(company_size) %>% 
  summarize(avg_salary = mean(salary_in_usd)) %>%
  ggplot()+ 
  geom_col(mapping=aes(x=company_size, y=avg_salary))+
  labs(x='Company Size', 
       y = 'Average Salary',
       title = 'Average Salary by the Company Size', 
       caption = 'This graph represents the average salary by the company size.')
```
```{r}
df %>%
  filter(job_title %in% c('Data Scientist', 'Data Analyst', 'Machine Learning Engineer', 'Research Scientist')) %>% 
  group_by(job_title) %>% 
  summarize(avg_salary = mean(salary_in_usd)) %>%
  ggplot()+ 
  geom_col(mapping=aes(x=job_title, y=avg_salary))+
  labs(x='Job Title', 
       y = 'Average Salary',
       title = 'Average Salary depending on Job Title', 
       caption = 'This graph represens the average salary for 4 of the top job titles with the highest average salary in USD.')
```


